---
title: "AnAge mammalian lifespan extraction"
author: "Alexander Tchernev"
date: "`r Sys.Date()`"
output: html_document
---

# Packages
```{r}
library(dplyr)
library(ggplot2)
```


# Mammalian longevity

## Load AnAge
```{r}
AnAge = read.delim("D:/Birmingham/Away Mini Project/AnAge/anage_data.txt")
```


## Extract mammals
```{r}
mammals = AnAge[AnAge$Class == "Mammalia", ]
```


## Remove species for which longevity data is not available
```{r}
mammalian_longevity = mammals[!is.na(mammals$Maximum.longevity..yrs.), ]
```


## Merge the genus and species columns into a new column and then remove all columns except for the newly merged column, the longevity column, the adult body mass column, the sample size column, and the data quality column
```{r}
# Merge columns
genus_and_species = paste0(mammalian_longevity$Genus, "_", mammalian_longevity$Species)

# Add longevity data
longevity = data.frame(Species = genus_and_species, 
							  Max_longevity = mammalian_longevity$Maximum.longevity..yrs., 
							  Adult_body_mass = mammalian_longevity$Adult.weight..g., 
							  Sample_size = mammalian_longevity$Sample.size, 
							  Data_quality = mammalian_longevity$Data.quality)
```


## Save the longevity dataframe as a tab-delimited file
```{r}
write.table(longevity, "D:/Birmingham/Away Mini Project/AnAge/longevity.txt", sep = "\t", dec = ".", row.names = F, col.names = F, quote = F)
```


## Save the species names
```{r}
write.table(genus_and_species, "D:/Birmingham/Away Mini Project/AnAge/names.txt", sep = "\t", dec = ".", row.names = F, col.names = F, quote = F)
```



# Primates longevity

## Extract only primates with longevity data
```{r}
primates_longevity = mammalian_longevity[mammalian_longevity$Order == "Primates", ]
```


## Merge the genus and species columns into a new column and then remove all columns except for the newly merged column, the longevity column, the sample size column, and the data quality column
```{r}
# Merge columns
genus_and_species = paste0(primates_longevity$Genus, "_", primates_longevity$Species)

# Add longevity data
primates_longevity = data.frame(Species = genus_and_species, 
										  Max_longevity = primates_longevity$Maximum.longevity..yrs., 
										  Adult_body_mass = primates_longevity$Adult.weight..g., 
										  Sample_size = primates_longevity$Sample.size, 
										  Data_quality = primates_longevity$Data.quality)
```



# RefSeq primates

## Filter the primates for which a RefSeq genome was found when running the perl script
```{r}
# Define a vector of the primates with genomes in RefSeq
refseq_primates = c("Homo_sapiens", "Hylobates_moloch", "Lemur_catta", "Macaca_fascicularis", 
						  "Macaca_mulatta", "Macaca_nemestrina", "Mandrillus_leucophaeus", 
						  "Microcebus_murinus", "Nomascus_leucogenys", "Nycticebus_coucang", 
						  "Otolemur_garnettii", "Pan_paniscus", "Pongo_pygmaeus", 
						  "Rhinopithecus_roxellana", "Sapajus_apella", "Symphalangus_syndactylus", 
						  "Theropithecus_gelada", "Trachypithecus_francoisi")

# Extract the records for those primates from the table for all primates
refseq_primates_long = filter(primates_longevity, primates_longevity$Species %in% refseq_primates)

# Add a column that states that all the species belong to primates (this has to be present to satisfy the grouping requirement of CALANGO) and a column with a normalization value of 1 for each species (this also has to be present to satisfy the normalization requirement of CALANGO)
refseq_primates_long = mutate(refseq_primates_long, Group = "primates", Norm = 1)

# Save the table
write.table(refseq_primates_long, "D:/Birmingham/Away Mini Project/AnAge/primates_metadata.txt", sep = "\t", dec = ".", row.names = F, col.names = F, quote = F)
```



# Save a data frame version where the longevity values are ln-transformed
```{r}
# Create a copy of the data frame
ln_refseq_primates_long = refseq_primates_long

# Ln-transform the longevity values
ln_refseq_primates_long$Max_longevity = log(ln_refseq_primates_long$Max_longevity)

# Save the data frame
write.table(ln_refseq_primates_long, "D:/Birmingham/Away Mini Project/AnAge/ln_primates_metadata.txt", sep = "\t", dec = ".", row.names = F, col.names = F, quote = F)

# Have a look at the normal Q-Q plot to judge normality of lifespan distribution
qqnorm(ln_refseq_primates_long$Max_longevity)
qqline(ln_refseq_primates_long$Max_longevity)
```


<!-- ## Save table version where the longevity values are log-transformed -->
<!-- ```{r} -->
<!-- lg_refseq_primates_long = refseq_primates_long -->

<!-- lg_refseq_primates_long$Max_longevity = log10(lg_refseq_primates_long$Max_longevity) -->

<!-- write.table(lg_refseq_primates_long, "D:/Birmingham/Away Mini Project/AnAge/lg_primates_metadata.txt", sep = "\t", dec = ".", row.names = F, col.names = F, quote = F) -->

<!-- # Have a look at the normal Q-Q plot to judge normality of lifespan distribution -->
<!-- qqnorm(lg_refseq_primates_long$Max_longevity) -->
<!-- qqline(lg_refseq_primates_long$Max_longevity) -->
<!-- ``` -->



# Plot max longevity values - raw and transformed - to see distribution
```{r}
# Longevity - not transformed
ggplot(refseq_primates_long, aes(x = Max_longevity)) +
	geom_histogram(aes(y = ..density..), binwidth = 4, fill = "blue", color = "black") +
	geom_density(alpha = 0.4, fill = "darkorange") +
	labs(title = "Histogram with Density", x = "Max longevity - years", y = "Density") +
	theme(axis.title = element_text(size = 14), 
			axis.title.y = element_text(margin = margin(r = 10)), 
			plot.title = element_text(hjust = 0.5, size = 16))

# Longevity - natural log transformed
ggplot(refseq_primates_long, aes(x = log(Max_longevity))) +
	geom_histogram(aes(y = ..density..), binwidth = 0.07, fill = "blue", color = "black") +
	geom_density(alpha = 0.4, fill = "darkorange") +
	labs(title = "Histogram with Density", x = "Max longevity - ln(years)", y = "Density") +
	theme(axis.title = element_text(size = 14), 
			axis.title.y = element_text(margin = margin(r = 10)), 
			plot.title = element_text(hjust = 0.5, size = 16))

# Longevity - log10 transformed
ggplot(refseq_primates_long, aes(x = log10(Max_longevity))) +
	geom_histogram(aes(y = ..density..), binwidth = 0.03, fill = "blue", color = "black") +
	geom_density(alpha = 0.4, fill = "darkorange") +
	labs(title = "Histogram with Density", x = "Max longevity - lg(years)", y = "Density") +
	theme(axis.title = element_text(size = 14), 
			axis.title.y = element_text(margin = margin(r = 10)), 
			plot.title = element_text(hjust = 0.5, size = 16))
```



# Add two columns to the data frame that contain the ln-transformed longevity and adult body mass values
```{r}
refseq_primates_long = refseq_primates_long |> mutate(ln_Max_longevity = log(Max_longevity), ln_Adult_body_mass = log(Adult_body_mass))
```



# With humans

## Linear regression and residuals extraction

### Plot the ln-transformed longevity values against the ln-transformed adult body mass values
```{r}
ggplot(refseq_primates_long, aes(x = ln_Adult_body_mass, y = ln_Max_longevity)) +
	geom_point() +
	geom_smooth(method = "lm") +
	labs(title = "Ln-transformed values", x = "Adult Body Mass", y = "Max Longevity") +
	theme(axis.title = element_text(size = 14), 
			axis.title.y = element_text(margin = margin(r = 10)), 
			plot.title = element_text(hjust = 0.5, size = 16))
```


### Linear regression
```{r}
# Model
Mass_vs_Long = lm(refseq_primates_long$ln_Max_longevity ~ refseq_primates_long$ln_Adult_body_mass)

# Anova table
anova(Mass_vs_Long)

# Model summary
summary(Mass_vs_Long)
```


### Residuals extraction
```{r}
# Residuals
resid = Mass_vs_Long$residuals

# Residuals distribution
hist(resid)

# Add column containing residuals from ln-adult body mass vs ln-max longevity to data frame that will be used as metadata file
Homo_resid_refseq_primates_long = mutate(refseq_primates_long, Resid = resid)

# Reorder the columns
Homo_resid_refseq_primates_long = Homo_resid_refseq_primates_long[, c(1, 10, 6, 7, 2, 8, 3, 9, 4, 5)]

# Save the data frame
write.table(Homo_resid_refseq_primates_long, 
				"D:/Birmingham/Away Mini Project/AnAge/Homo_resid_ln_primates_metadata.txt", 
				sep = "\t", dec = ".", row.names = F, col.names = F, quote = F)
```


# Without humans

## Remove humans because they are outliers
```{r}
no_Homo_refseq_primates_long = filter(refseq_primates_long, Species != "Homo_sapiens")
```

## Linear regression and residuals extraction

### Plot the ln-transformed longevity values against the ln-transformed adult body mass values
```{r}
ggplot(no_Homo_refseq_primates_long, aes(x = ln_Adult_body_mass, y = ln_Max_longevity)) +
	geom_point() +
	geom_smooth(method = "lm") +
	labs(title = "Ln-transformed values", x = "Adult Body Mass", y = "Max Longevity") +
	theme(axis.title = element_text(size = 14), 
			axis.title.y = element_text(margin = margin(r = 10)), 
			plot.title = element_text(hjust = 0.5, size = 16))
```



### Linear regression
```{r}
# Model
no_Homo_Mass_vs_Long = lm(no_Homo_refseq_primates_long$ln_Max_longevity ~ no_Homo_refseq_primates_long$ln_Adult_body_mass)

# Anova table
anova(no_Homo_Mass_vs_Long)

# Model summary
summary(no_Homo_Mass_vs_Long)
```


### Residuals extraction
```{r}
# Residuals
no_Homo_resid = no_Homo_Mass_vs_Long$residuals

# Residuals distribution
hist(no_Homo_resid)

# Add column containing residuals from ln-adult body mass vs ln-max longevity to data frame that will be used as metadata file
no_Homo_resid_refseq_primates_long = mutate(no_Homo_refseq_primates_long, Resid = no_Homo_resid)

# Reorder the columns
no_Homo_resid_refseq_primates_long = no_Homo_resid_refseq_primates_long[, c(1, 10, 6, 7, 2, 8, 3, 9, 4, 5)]

# Save the data frame
write.table(no_Homo_resid_refseq_primates_long, 
				"D:/Birmingham/Away Mini Project/AnAge/no_Homo_resid_ln_primates_metadata.txt", 
				sep = "\t", dec = ".", row.names = F, col.names = F, quote = F)
```